\emph{icepack}: a novel glacier flow modeling package

### 2019-03-24 icepack.tex
Wrote some about prognostic solver

One of our main design goals for icepack is to enable users to ``plug and play'' their own physics.
Since the physics of glacier flow are not understood completely, this focus on extensibility is essential for glaciological modeling \textcolor{red}{reword somehow}.
Writing the entire package in Python was necessary to meet this goal.
Many scientific software packages consist of a highly-optimized core library in C, C++, or Fortran, together with a wrapper that exposes this library to an interpreted language like Python.
This approach in principle offers the best of both worlds -- the runtime efficiency of a compiled language with the convenience and interactivity of an interpreted one.
Many practitioners, however, will use such a library entirely through the interpreted surface layer and have no familiarity with the language in which the fast core was written.
These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
Some low-level kernel code is always necessary for runtime efficiency of course.
With Firedrake, this responsibility is deferred to a code generator of which the average user remains blissfully unaware.

@@

>>>>

One of our main design goals for icepack is to enable users to ``plug and play'' their own physics.
Since many aspects of glacier flow are not understood completely, we view extensibility as an essential requirement for any software package in this domain.
Writing the entire package in Python was necessary to meet this goal.
Many scientific software packages consist of a highly-optimized core library in C, C++, or Fortran, together with a wrapper that exposes this library to an interpreted language like Python.
This approach in principle offers the best of both worlds -- the runtime efficiency of a compiled language with the convenience and interactivity of an interpreted one.
Many practitioners, however, will use such a library entirely through the interpreted surface layer and have no familiarity with the language in which the fast core was written.
These users then have a steep learning curve to overcome when they need to modify functionality that lives in the core library, as they now need to learn an entirely new programming language.
Some low-level kernel code is always necessary for runtime efficiency of course.
With Firedrake, this responsibility is deferred to a code generator of which the average user remains blissfully unaware.

@@

The PDE describing how the ice thickness changes in time is a simple conservative advection equation:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
The thickness is represented using continuous Galerkin basis functions.
The ice thickness and bed elevation are used to update the surface elevation at every timestep, and we need the surface gradient to calculate the driving stress.
The need to calculate gradients of the surface elevation precludes the use of discontinuous Galerkin methods.

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
Other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \textcolor{red}{cite ISSM, VarGlaS}.
The SUPG method is explicit in time and conditionally stable \textcolor{red}{cite Donea and Huerta?}.
This scheme introduces a tuneable stabilization parameter, so in the interest of simplicity we instead default to the implicit Euler scheme.
The implicit Euler scheme requires solving a more complicated linear system, but the extra computational cost comes with the advantage of unconditional stability.
Practicing glaciologists who have not studied numerical PDE may be unfamiliar with the Courant-Friedrichs-Lewy condition.
The use of an unconditionally stable scheme guarantees that they will get an answer should they try to use a large timestep rather than a runtime error.
The extra cost of using an implicit time discretization for the mass transport equation is dwarfed by the cost of the diagnostic solve in any case.
Advanced users who are interested in maximizing performance can subclass the mass transport solver to implement a faster explicit scheme.

The implicit Euler scheme introduces a large degree of artificial diffusion \textcolor{red}{cite Donea and Huerta}.
Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.
Other problems within glaciology have more of a hyperbolic character and there the argument for using shock-capturing methods is much stronger.
For example, in continuum damage mechanics, the thresholding behavior of the source terms can induce sharp discontinuities in the damage field \textcolor{red}{cite Albrecht and Levermann}.
The implicit Euler scheme would obscure this feature and an alternative method such as the strong stability-preserving Runge Kutta method would be more appropriate.
